---
title: "Non-Linear WQS for Mixture Exposures: an applied walkthrough"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Non-Linear WQS for Mixture Exposures: an applied walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4,
  dpi    = 96,
  warning = FALSE
)
set.seed(2026)
```

```{r load}
library(NWQS)
```

## When to use NWQS

Environmental-health studies routinely have to summarize the joint effect of a
*set* of correlated exposures — urinary metals, phthalate metabolites, PFAS
analytes, indoor air pollutants — on a health endpoint. Classical weighted
quantile sum (WQS) regression collapses the mixture into a single index. NWQS
extends WQS in two practical ways:

1. each component is expanded into a natural cubic spline basis on the
   percentile-rank scale, so a non-linear dose-response (threshold, U-shape,
   S-shape) is preserved instead of being forced into a single linear term;
2. component weights are learned by **out-of-bag permutation importance**
   inside repeated holdout splits, which avoids the regularization bias of
   constrained-optimization WQS.

The package supports `gaussian`, `binomial`, `poisson`, `quasipoisson`, and
`negbin` outcomes. This vignette walks through a small simulated study from
end to end.

## A simulated metals → lung function cohort

We construct a cohort of $n = 400$ adults with seven correlated urinary metal
biomarkers (As, Cd, Pb, Hg, Mn, Ni, Se), age, and sex. The "true" mixture
effect is non-linear: As and Cd drive a meaningful drop in lung function once
their exposure exceeds the population median, while the other five metals are
essentially noise.

```{r data}
gen_metals_cohort <- function(n = 400, seed = 11) {
  set.seed(seed)
  p <- 7
  metals <- c("As", "Cd", "Pb", "Hg", "Mn", "Ni", "Se")
  rho <- 0.45
  Sigma <- matrix(rho, p, p); diag(Sigma) <- 1
  L <- chol(Sigma)
  X <- matrix(rnorm(n * p), n, p) %*% L
  colnames(X) <- metals
  # Percentile-rank to mimic biomarker scaling, then a non-linear effect on As/Cd
  U <- apply(X, 2, function(col) rank(col, ties.method = "average") / n)
  shape <- function(u) pmin(u, 0.5) * 0 + pmax(u - 0.5, 0) * (-2)
  eta_mix <- shape(U[, "As"]) * 0.7 + shape(U[, "Cd"]) * 0.5
  age <- rnorm(n, mean = 50, sd = 12)
  sex <- factor(sample(c("F", "M"), n, replace = TRUE))
  fev1 <- 3.5 - 0.012 * (age - 50) + ifelse(sex == "M", 0.45, 0) +
          eta_mix + rnorm(n, sd = 0.30)
  data.frame(fev1 = fev1, as.data.frame(X), age = age, sex = sex)
}
cohort <- gen_metals_cohort()
head(cohort, 3)
```

## Fitting an NWQS model

`nwqs()` does the heavy lifting in one call. The 0.2.0 defaults are
`transform_type = "percentile_rank"`, `n_permutation = 30`, `rh = 10`. For the
vignette we set `rh = 5` so the chunk runs quickly.

```{r fit}
metals <- c("As", "Cd", "Pb", "Hg", "Mn", "Ni", "Se")

fit <- nwqs(
  data           = cohort,
  mix_name       = metals,
  covariates     = c("age", "sex"),
  outcome        = "fev1",
  family         = "gaussian",
  transform_type = "percentile_rank",
  q              = 4,
  rh             = 5,
  n_permutation  = 5,
  seed           = 1234,
  quiet          = TRUE
)
print(fit)
```

The `final_weights` field gives each component's share of the mixture index;
As and Cd should dominate by construction.

```{r weights}
round(sort(fit$final_weights, decreasing = TRUE), 3)
```

`plot(fit)` produces dose-response and weight diagnostics in one shot.

```{r plot, fig.alt = "Dose-response curves and weight bars for the NWQS fit"}
plot(fit)
```

## Valid sampling-variance inference: `nwqs_boot()`

`fit$fit$coefficients` reports standard errors derived from the repeated
holdout splits. Those are **algorithmic variance** — the spread across data
splits with the original sample held fixed. They are not valid for inference.
Use `nwqs_boot()` to get a real percentile bootstrap.

```{r boot}
boot_fit <- nwqs_boot(
  data           = cohort,
  mix_name       = metals,
  covariates     = c("age", "sex"),
  outcome        = "fev1",
  family         = "gaussian",
  transform_type = "percentile_rank",
  q              = 4,
  n_boot         = 30,
  rh_inner       = 1,
  n_permutation  = 5,
  seed           = 1234,
  quiet          = TRUE
)
print(boot_fit)
```

Bootstrap percentile CIs for every term are in `boot_fit$ci_table`.

```{r boot-ci}
boot_fit$ci_table
```

`confint(boot_fit)` returns the same intervals in `confint.glm()`-compatible
matrix form, ready for table assembly.

```{r boot-confint}
confint(boot_fit)
```

## Predictions on new data

The 0.2.0 `predict.nwqs()` method scores new observations on the **training**
empirical distribution, so there is no train/predict drift. `type` selects the
scale: `"nwqs_index"`, `"link"`, or `"response"` (the default).

```{r predict}
newdat <- gen_metals_cohort(n = 30, seed = 99)
preds <- predict(fit, newdata = newdat, type = "response")
head(data.frame(observed = newdat$fev1, predicted = preds), 5)
```

## Reporting with `broom`

If `broom` (and `generics`) are installed, NWQS dispatches through the broom
S3 generics automatically. `tidy(fit, conf.int = TRUE)` is publication-ready.

```{r broom, eval = requireNamespace("broom", quietly = TRUE)}
broom::tidy(boot_fit)
broom::glance(boot_fit)
```

## A count-outcome example with `family = "negbin"`

Many environmental endpoints are overdispersed counts — visits to a clinic,
days with reported symptoms. `family = "negbin"` dispatches the final
regression to `MASS::glm.nb()`; the rest of the pipeline (transforms,
permutation importance, bootstrap CIs, predict, confint, tidy) is unchanged.

```{r negbin}
nb_dat <- gen_nbin_data(
  n_obs     = 400,
  n_vars    = 5,
  beta      = c(0.7, -0.4, 0.0, 0.0, 0.0),
  intercept = 1.0,
  theta     = 2.0,
  snr_db    = 15,
  seed      = 7
)

nb_fit <- nwqs(
  data           = nb_dat,
  mix_name       = paste0("V", 1:5),
  outcome        = "y",
  family         = "negbin",
  rh             = 5,
  n_permutation  = 5,
  seed           = 1234,
  quiet          = TRUE
)
round(sort(nb_fit$final_weights, decreasing = TRUE), 3)
```

## Reproducing the 0.1.x defaults

If a published analysis used 0.1.x, you can reproduce its discrete-quartile
transform by passing the explicit opt-in arguments:

```{r legacy, eval = FALSE}
nwqs(
  ...,
  transform_type = "q_bin",
  q              = 4
)
```

See `NEWS.md` for the full list of breaking changes between 0.1.x and 0.2.0
and for the migration recipes.

## Where to go next

- `?nwqs` for the full parameter list.
- `?nwqs_boot` for bootstrap-specific options.
- `?nwqs_control` for advanced knobs (`custom_knots`, `custom_boundary`).
- `tests/testthat/` in the source tree for runnable contract examples of
  every public method.
